A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-17 tex/codatadp.tex
more

\end{report}
A related semantic view of datatypes is as initial algebras \cite{xxx}.
Codatatypes are then defined dually as final coalgebras \cite{xxx}. The
datatypes are generated by their constructors, whereas the codatatypes are
viewed through their selectors. By uniformly focusing on the constructors, the
axiomatic approach emphasizes the commonality between datatypes and
codatatypes, while sacrificing a theoretically fruitful notion of duality.

@@

\paragraph{Special Cases.}
A type $\delta$ depends on another type $\varepsilon$ if $\varepsilon$ is the
type of an argument to one of $\delta$'s constructors. Semantically, a set of
types is mutually (co)recursive if and only if the associated dependency graph
is strongly connected. Types can be declared together as mutually
(co)recursive even if they are not actually (co)recursive; however, the
semantic notion is more precise and is the one that interests us.
%
In addition, nothing forbids non(co)recursive specifications
such~as
\begin{paper}
\[\keyw{datatype}\;\, \ty{option}_{\,\tau} \,=\, \const{None} \,\mid\, \const{Some}(\tau)\]%
\end{paper}%
\begin{report}%
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{option}_{\,\tau} & \,=\, \const{None} \mid \const{Some}(\tau) \\[-.5\jot]
      \keyw{codatatype}\; \,\ty{complex} & \,=\, \const{Complex}(\const{re}{:}\; \ty{real},\; \const{im}{:}\;\ty{real})
\end{aligned}
\]
At the semantic level, it makes no difference whether such types are
introduced as datatypes or as codatatypes.
\end{report}%
%Without loss of generality, we consider that these types are datatypes
%instead of codatatypes.

Some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\vthinspace$\keyw{codatatype}~\ty{a} = \const{A}(\ty{a})$, whose
only value is $\mu a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-1\jot]
      \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$, as well as $\ty{stream}_{\,\ty{unit}\,}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
important to detect these types. %, even if they rarely arise in practice.
A type may also be a corecursive singleton only in some models. If the example
above is altered to leave \ty{unit} uninterpreted, \ty{b} and \ty{c} will be
singletons precisely when \ty{unit} is interpreted as a singleton.
Fortunately, given cardinalities for the ordinary types, it is easy to
characterize this degenerate case:

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec,\: t \teq t
}
\)
\rn{Refl}
\qquad
\(
\inferrule{
 t_1 \teq t_2 \in \Ec
}{
 \Ec := \Ec,\: t_2 \teq t_1
}
\)
\rn{Sym}
\qquad
\(
\inferrule{
  t_1 \teq t_2,\; t_1 \tneq t_2 \in \Ec
}{
  \bot
}
\)
\rn{Conflict}
\\[5\jot]
\(
\inferrule{
  t_1 \teq t_2,\; t_2 \teq t_3 \in \Ec
}{
  \Ec := \Ec,\: t_1 \teq t_3
}
\)
\rn{Trans}
\qquad
\(
\inferrule{
  \vec t \teq \vec u \in \Ec \quad \const f( \vec t \,),\, \const f( \vec u ) \in \tEc
}{
  \Ec := \Ec,\: \const f( \vec t \,) \teq \const f( \vec u )
}
\)
\rn{Cong} 
\\[5\jot]
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{C}( \vec u ) \in \Ec
}{
  \Ec := \Ec,\: \vec t \teq \vec u
}
\)
\rn{Decomp} 
\qquad
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{D}( \vec u ) \in \Ec
  \quad
  \const{C} \not= \const{D}
}{
  \bot
}
\)
\rn{Clash}
\end{tabular}
\caption{\,Rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\(
\inferrule{
  \tau \in \Data
  \quad
  t_1 : \tau
  \quad
  %\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset
  \Val \ec{t_1} = \mu x.\; t_2
  \quad
  x \in FV( t_2 )
}{
  \bot
}
\)
\rn{Cyclic}
\\[4\jot]
\(
\inferrule{
 \tau \in \Codata
 \quad
 t_1, t_2 : \tau
 \quad
 %\tpath{\Ec}{\ec{t_1}} = \tpath{\Ec}{\ec{t_2}}\sigma \neq \emptyset
 \Val \ec{t_1} \vsim \Val \ec{t_2}
}{
 \Ec := \Ec,\: t_1 \teq t_2
}
\)
\rn{Bisimilar}
\end{tabular}
\caption{\,Rules for cyclicity and bisimilarity%.
}
\label{fig:ab-rules}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\(
\inferrule{
  t, u : \tau \in \tEc
  \quad
  \tau \in \Codata,
  \mid \tau \mid = 1
}{
  \Ec := \Ec, t \teq u
}
\)
\rn{Singleton}
\\[5\jot]
\(
\inferrule{
  t : \tau \in \tEc 
  \quad 
  ( s( t ) \in \tEc, s \in \Sel^\tau ) 
  \text{ or } 
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec, t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \ldots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\rn{Split} 
\end{tabular}
\caption{\,Rules for branching%.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

>>>>

\end{report}
A related semantic view of datatypes is as initial algebras \cite{xxx}.
Codatatypes are then defined dually as final coalgebras \cite{xxx}. The
datatypes are generated by their constructors, whereas the codatatypes are
viewed through their selectors.
%By uniformly focusing on the constructors, the
%axiomatic approach emphasizes the commonality between datatypes and
%codatatypes, while sacrificing a theoretically fruitful notion of duality.

@@

\paragraph{Special Cases.}
A type $\delta$ depends on another type $\varepsilon$ if $\varepsilon$ is the
type of an argument to one of $\delta$'s constructors. Semantically, a set of
types is mutually (co)recursive if and only if the associated dependency graph
is strongly connected. Types can be declared together as mutually
(co)recursive even if they are not actually (co)recursive; however, the
semantic notion is more precise and is the one that interests us.
%
In addition, nothing forbids non(co)recursive specifications
such~as
\begin{paper}%
\vthinspace$\keyw{datatype}~\vthinspace\ty{option}_{\,\tau} = \const{None} \mid \const{Some}(\tau)$.%
\end{paper}%
\begin{report}%
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{option}_{\,\tau} & \,=\, \const{None} \mid \const{Some}(\tau) \\[-.5\jot]
      \keyw{codatatype}\; \,\ty{complex} & \,=\, \const{Complex}(\const{re}{:}\; \ty{real},\; \const{im}{:}\;\ty{real})
\end{aligned}
\]
At the semantic level, it makes no difference whether such types are
introduced as datatypes or as codatatypes.
\end{report}%
%Without loss of generality, we consider that these types are datatypes
%instead of codatatypes.

Some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\vthinspace$\keyw{codatatype}~\ty{a} = \const{A}(\ty{a})$, whose
only value is $\mu a.\; A(a)$. Other specimens are
\begin{paper}%
\vthinspace$\keyw{codatatype}~\ty{b} = \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit})
~\keyw{and}~ \ty{c} = \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})$,
\end{paper}%
\begin{report}%
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-1\jot]
      \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
\end{report}
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$, as well as $\ty{stream}_{\,\ty{unit}\,}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
important to detect these types. %, even if they rarely arise in practice.
A type may also be a corecursive singleton only in some models. If the example
above is altered to leave \ty{unit} uninterpreted, \ty{b} and \ty{c} will be
singletons precisely when \ty{unit} is interpreted as a singleton.
Fortunately, given cardinalities for the ordinary types, it is easy to
characterize this degenerate case:

@@

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec,\: t \teq t
}
\)
\rn{Refl}
\qquad
\(
\inferrule{
 t_1 \teq t_2 \in \Ec
}{
 \Ec := \Ec,\: t_2 \teq t_1
}
\)
\rn{Sym}
\qquad
\(
\inferrule{
  t_1 \teq t_2,\; t_1 \tneq t_2 \in \Ec
}{
  \bot
}
\)
\rn{Conflict}
\\[5\jot]
\(
\inferrule{
  t_1 \teq t_2,\; t_2 \teq t_3 \in \Ec
}{
  \Ec := \Ec,\: t_1 \teq t_3
}
\)
\rn{Trans}
\qquad
\(
\inferrule{
  \vec t \teq \vec u \in \Ec \quad \const f( \vec t \,),\, \const f( \vec u ) \in \tEc
}{
  \Ec := \Ec,\: \const f( \vec t \,) \teq \const f( \vec u )
}
\)
\rn{Cong} 
\\[5\jot]
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{C}( \vec u ) \in \Ec
}{
  \Ec := \Ec,\: \vec t \teq \vec u
}
\)
\rn{Decomp} 
\qquad
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{D}( \vec u ) \in \Ec
  \quad
  \const{C} \not= \const{D}
}{
  \bot
}
\)
\rn{Clash}
\end{tabular}
\caption{\,Rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

@@

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  \tau \in \Data
  \quad
  t_1 : \tau
  \quad
  %\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset
  \Val \ec{t_1} = \mu x.\; t_2
  \quad
  x \in FV( t_2 )
}{
  \bot
}
\)
\rn{Cyclic}
\\[4\jot]
\(
\inferrule{
 \tau \in \Codata
 \quad
 t_1, t_2 : \tau
 \quad
 %\tpath{\Ec}{\ec{t_1}} = \tpath{\Ec}{\ec{t_2}}\sigma \neq \emptyset
 \Val \ec{t_1} \vsim \Val \ec{t_2}
}{
 \Ec := \Ec,\: t_1 \teq t_2
}
\)
\rn{Bisimilar}
\end{tabular}
\caption{\,Rules for cyclicity and bisimilarity%.
}
\label{fig:ab-rules}
\end{figure}

@@

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  t, u : \tau \in \tEc
  \quad
  \tau \in \Codata,
  \mid \tau \mid = 1
}{
  \Ec := \Ec, t \teq u
}
\)
\rn{Singleton}
\\[5\jot]
\(
\inferrule{
  t : \tau \in \tEc 
  \quad 
  ( s( t ) \in \tEc, s \in \Sel^\tau ) 
  \text{ or } 
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec, t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \ldots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\rn{Split} 
\end{tabular}
\caption{\,Rules for branching%.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}
